---
title: "Assignment 1 - Data wrangling and functions"
author: "YENUS IBRAHIM AYALEW"
username: "YenusAyalew"
date: "`r format(Sys.time(), '%B %d, %Y | %H:%M:%S | %Z')`"
output:
  pdf_document:
    latex_engine: xelatex   # Needed for emojis like 🚫 💪
  html_document:
    code_folding: show      # Allows collapsing code in HTML
    df_print: paged         # Pretty tables for data frames
    highlight: tango        # Code highlighting style
    number_sections: false
    theme: cosmo            # HTML theme
    toc: false
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<style>
div.answer {background-color:#f3f0ff; border-radius: 5px; padding: 20px;}
div.comment {background-color:#F0F6FF; border-radius: 5px; padding: 20px;}

</style>
```

```{=html}
<style>
div.gradingadvice {background-color:#a2bc89; border-radius: 5px; padding: 20px;}
</style>
```

```{=html}
<style>
div.examplefeedback {background-color:#FFA500; border-radius: 5px; padding: 20px;}
</style>
```

```{=html}
<style>
div.lookfor {background-color:#a389bc; border-radius: 5px; padding: 20px;}
</style>
```

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = TRUE,
                      error = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      comment = NA)
```

<!-- Do not forget to input your Github username in the YAML configuration up there -->

------------------------------------------------------------------------

```{r, include = T}
# LOAD THE PACKAGES YOU ARE USING IN THIS CODE CHUNK
# library(nameofpackage)

library(tidyverse)

```

<br>

## Task 1: Getting started with the Astronaut Database

The data this week comes from [Mariya Stavnichuk and Tatsuya
Corlett](https://data.mendeley.com/datasets/86tsnnbv2w/1).

This
[article](https://www.sciencedirect.com/science/article/abs/pii/S2214552420300444)
talks about the data set in greater detail.

> This database contains publically available information about all
> astronauts who participated in space missions before 15 January 2020
> collected from NASA, Roscosmos, and fun-made websites. The provided
> information includes full astronaut name, sex, date of birth,
> nationality, military status, a title and year of a selection program,
> and information about each mission completed by a particular astronaut
> such as a year, ascend and descend shuttle names, mission and
> extravehicular activity (EVAs) durations.

Credit for preparing the data set: [Georgios
Karamanis](https://twitter.com/geokaramanis)

You can use the data set directly by importing the csv file in the
repository.

------------------------------------------------------------------------

If not noted otherwise, for the following tasks, you will work only with
a subset of astronauts based on their **year of birth**. The year span
you are assigned depends on your first name:

| Your first name starts with... | Years of birth |
|--------------------------------|----------------|
| A-I                            | (1921-1942]    |
| J-R                            | (1943-1964]    |
| S-Z                            | (1965-1983]    |

since my name starts with "Y"

1965 to 1983 will be my assigned year <br>

------------------------------------------------------------------------

a)  How many unique astronauts were there in the full data set and how
    many are there in your subset? (*Note that some might have been part
    of multiple missions.*)

```{r}
# Unique astronuts in the full data set and in my subset
# load the data first
astronuts_data <- read.csv("astronauts.csv")
# column names of the data 
colnames(astronuts_data)

# dplyr loading 
library(dplyr)
#Unique number of astronauts
astronuts_data %>% 
  distinct(name) %>% 
  nrow() # This is total number of unique astronuts
#  For subset give to me
astronuts_data %>% 
  filter(year_of_birth >= 1965 & year_of_birth <= 1983) %>%
  distinct(name) %>% 
  nrow() # The number of unique astronauts with birth year of 1965-1983
  
```

b)  What's the overall share of female astronauts in the subset of data
    you were assigned to? And what's the share of astronauts with U.S.
    nationality?

```{r}
# Overall share of female astronauts for subset I have been assigned
my_subset <- astronuts_data %>% 
  filter(year_of_birth >= 1965 & year_of_birth <= 1983)
# Calculate female share
female_share <- my_subset %>%
  summarise(
    total_astronauts = n(),
    female_count = sum(sex == "female", na.rm = TRUE),
    female_share = (female_count / total_astronauts) * 100
  )
cat("Share of female astronauts:", round(female_share$female_share, 2), "%\n")
Us_nationality <- my_subset %>% 
  summarise(
    total_astronauts = n(),
    US_national_count = sum(nationality == "U.S."),  
    US_share = (US_national_count/total_astronauts) * 100
  )
cat("Share of US Nationals:", round(Us_nationality$US_share, 2), "%\n")
```

<br> c) Create a new data set `total_hours_df` containing the unique
names and the total duration of all missions in hours (`total_hrs_sum`)
for each astronaut in the subset. Then, visualize the distribution of
that variable using an appropriate plot type.

```{r}
# new data set 
total_hours_df <- my_subset %>%
  group_by(name) %>%
  summarise(total_hrs_sum = sum(hours_mission, na.rm = TRUE)) %>%
  ungroup()
head(total_hours_df) #


# Since it is continous data we can use histogram to plot it
hist(total_hours_df$total_hrs_sum, xlab = "Hours of mission", main = "Distribution of the total duration\nof all missions in hours for each astronaut")


```

<br> d) Using the full data set, provide a plot that shows the overall
share of *civilian* astronauts by the year of mission!

```{r}
## 
civilian_share <- astronuts_data %>%
  group_by(year_of_mission) %>%
  summarise(civilian_share = mean(military_civilian == "civilian", na.rm = TRUE) * 100)
# Create plot: since it shows trend line graph is more appropriate 
ggplot(civilian_share, aes(x = year_of_mission, y = civilian_share)) +
  geom_line() +
  labs(x = "Mission Year", y = "Civilian Share (%)")

```

<br> ***\### Task 2 – Silly Password Strength Checker Program a function
`check_password(pw)` that takes a password string `pw` and returns a
funny*** **strength rating\* according to the following rules:\
1. If the password contains the word `"password"` (in any mix of
upper/lower case), return `"Terrible! 🚫"`.\
2. If the password is shorter than 6 characters, return
`"Too weak 💤"`.\
3. If it contains only letters, return `"Meh 😐"`.\
4. If it contains letters** and\*\* numbers, return `"Better 👍"`.\
5. If it contains letters **and** numbers **and** at least one special
character from `!@#$%^&*?`, return `"Strong 💪"`.

Write the function and test it with at least **five different
passwords** that trigger different outputs. Also, use functions from the
`stringr` package to execute the checks in your function (hint: look up
`str_detect()` and `str_length()`).

```{r}
library(stringr)

check_password <- function(pw) {
  # Rule 1: Contains "password" (case-insensitive)
  if (str_detect(pw, regex("password", ignore_case = TRUE))) {
    return("Terrible! 🚫")
  }
  
  # Rule 2: Too short
  if (str_length(pw) < 6) {
    return("Too weak 💤")
  }
  
  # Character class checks
  has_letter  <- str_detect(pw, "[A-Za-z]")
  has_number  <- str_detect(pw, "[0-9]")
  has_special <- str_detect(pw, "[!@#$%^&*?]")
  
  # Rule 5: letters + numbers + special
  if (has_letter && has_number && has_special) {
    return("Strong 💪")
  }
  
  # Rule 4: letters + numbers
  if (has_letter && has_number) {
    return("Better 👍")
  }
  
  # Rule 3: only letters
  if (str_detect(pw, "^[A-Za-z]+$")) {
    return("Meh 😐")
  }
  
  return("Unclear 🤔")
}

# Test cases
check_password("hello")        # Too weak 💤
check_password("hello123")     # Better 👍
check_password("PassWord!")    # Terrible! 🚫
check_password("abc!23")       # Strong 💪
check_password("abcdef")       # Meh 😐




```

<br> c) **Bonus challenge 2 (you're REALLY bored, right?):** Generalize
the Monty Hall Problem to k doors and k-1 goats, where the host will
open all doors except for 2. Now, visualize the probability of winning
(proportion of wins) for both strategies (switch and stay) as a function
of the number of simulated rounds n and the number of doors k, for a set
of values for both n and k that you choose. *Important:* This is just
for fun. You don't have to work on this task if you don't have time.
Providing a right/wrong solution does not give/cost you extra points. (A
right solution gives you kudos from the instructor though!)

```{r}
# ------------------------------------------------------------
# Task 3 – The Monty Hall Machine
# ------------------------------------------------------------
# This task simulates the classic Monty Hall problem (3 doors)
# and also generalizes it to k doors.
# The goal is to compare the probability of winning when
# staying vs switching doors.
# ------------------------------------------------------------

library(tidyverse)

# ------------------------------
# 1️⃣ Classic 3-door Monty Hall
# ------------------------------
monty_hall_choice <- function(switch = TRUE) {
  doors <- c(1, 2, 3)
  car_door <- sample(doors, 1)              # Randomly hide the car
  contestant_choice <- sample(doors, 1)     # Contestant chooses a door
  
  # Monty reveals a goat door (not car, not contestant choice)
  available_doors_for_monty <- setdiff(doors, c(contestant_choice, car_door))
  if (contestant_choice == car_door) {
    monty_choice <- sample(available_doors_for_monty, 1)
  } else {
    monty_choice <- available_doors_for_monty
  }
  
  # If switching, pick the remaining unopened door
  if (switch) {
    contestant_choice <- setdiff(doors, c(contestant_choice, monty_choice))
  }
  
  # Return TRUE if contestant wins
  contestant_choice == car_door
}

# Example simulation: 100 rounds
set.seed(123)
res_switch <- replicate(100, monty_hall_choice(switch = TRUE))
res_stay   <- replicate(100, monty_hall_choice(switch = FALSE))

mean(res_switch)  # Probability of winning if switching
mean(res_stay)    # Probability of winning if staying

# ------------------------------
# 2️⃣ Simulation wrapper for n rounds
# ------------------------------
monty_hall_simulation <- function(n = 1000, switch = TRUE) {
  wins <- sum(replicate(n, monty_hall_choice(switch = switch)))
  wins / n
}

# Test wrapper function
monty_hall_simulation(n = 10000, switch = TRUE)   # ~0.66
monty_hall_simulation(n = 10000, switch = FALSE)  # ~0.33

# ------------------------------
# 3️⃣ Generalized Monty Hall (k doors)
# ------------------------------
monty_hall_general <- function(k = 3, switch = TRUE) {
  if (k < 3) stop("k must be >= 3 because the host leaves 2 doors closed.")
  
  doors <- 1:k
  car_door <- sample(doors, 1)
  contestant_choice <- sample(doors, 1)
  
  # Doors Monty can open (not contestant's choice, not car)
  available_to_open <- setdiff(doors, c(contestant_choice, car_door))
  
  # Monty keeps exactly one other door closed in addition to contestant's choice
  if (contestant_choice == car_door) {
    monty_keep <- sample(available_to_open, 1)
  } else {
    monty_keep <- car_door
  }
  
  final_doors <- c(contestant_choice, monty_keep)
  
  # Switch or stay
  if (switch) {
    contestant_choice <- setdiff(final_doors, contestant_choice)
  }
  
  # Return TRUE if contestant wins
  contestant_choice == car_door
}

# Wrapper for generalized simulation
monty_hall_general_simulation <- function(n = 1000, k = 3, switch = TRUE) {
  wins <- sum(replicate(n, monty_hall_general(k = k, switch = switch)))
  wins / n
}

# ------------------------------
# 4️⃣ Run experiments and visualize
# ------------------------------
set.seed(123)
k_values <- c(3, 5, 10, 20, 100)
n_values <- c(100, 1000, 5000)

# Run simulations for all combinations of k, n, switch
results <- expand.grid(k = k_values, n = n_values, switch = c(FALSE, TRUE)) %>%
  mutate(win_prob = mapply(
    function(kk, nn, sw) monty_hall_general_simulation(n = nn, k = kk, switch = sw),
    k, n, switch
  ))

# Analytical probabilities for overlay
theory <- expand.grid(k = k_values, switch = c(FALSE, TRUE), n = n_values) %>%
  mutate(win_prob = ifelse(switch, (k - 1) / k, 1 / k))

# Smooth trend for wider k range
theory_smooth <- expand.grid(k = 3:50, switch = c(FALSE, TRUE), n = n_values) %>%
  mutate(win_prob = ifelse(switch, (k - 1) / k, 1 / k))

# Plot: Simulation vs Analytical
ggplot(results, aes(x = k, y = win_prob, color = switch, group = switch)) +
  geom_line(size = 1.2) +
  geom_point(position = position_dodge(width = 0.2), size = 2) +
  facet_wrap(~ n, labeller = labeller(n = function(x) paste("n =", x))) +
  
  geom_line(data = theory, aes(x = k, y = win_prob, color = switch, group = switch),
            linetype = "dashed", linewidth = 1.2) +
  geom_point(data = theory, aes(x = k, y = win_prob, color = switch),
             shape = 4, stroke = 1.1, size = 2.2) +
  
  geom_line(data = theory_smooth, aes(x = k, y = win_prob, color = switch, group = switch),
            linetype = "dotted", linewidth = 0.9) +
  
  labs(
    title = "Generalized Monty Hall (k doors, host leaves 2 doors closed)",
    subtitle = "Simulation (solid) vs analytical probabilities (dashed = chosen k; dotted = trend)",
    x = "Number of doors (k)",
    y = "Winning probability",
    color = "Strategy"
  ) +
  scale_color_manual(
    breaks  = c(FALSE, TRUE),
    labels  = c("Stay", "Switch"),
    values  = c("red", "blue")
  ) +
  scale_x_continuous(breaks = k_values) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")



```
